Report on the financial and human cost of weather events
========================================================

In this document we shall examine the leading causes of fatalities, injuries, and financial loss taken from the NOAA storm database.
We shall only consider years since 1970, since the early years of the database are more likely to be incomplete.
In each case, we shall report on the most costly (in financial or human terms) weather events.

```{r setoptions, echo = FALSE}
  opts_chunk$set(echo=FALSE, results="asis")
```
## Data Processing

All operations in the section are cached to save time on subsequent runs.
You should expect the first compilation of this document to spend considerable time on this section.

First we read in the data as a CSV file
```{r, cache = TRUE}
  workingDir <- getwd() 
  zipData <- paste(workingDir, "repdata%2Fdata%2FStormData.csv.bz2", sep="/")
  eventData <- read.csv(bzfile(zipData))
```

The **BGN_DATE** varaible is converted into the **POSIXlt** format.
```{r, cache = TRUE}
  # Experimentation with the dataset revealed that the
  # dates are in the month/day/year format
  eventData$PTIME = strptime(eventData$BGN_DATE, "%m/%d/%Y %H:%M:%S")
```

We make a new data frame containing only the variable we are interested in.
We summarise the data as a logical indicating whether or not the event was pre-1970.
```{r, cache = TRUE}
  costData = data.frame(EVTYPE = eventData$EVTYPE, FATALITIES = eventData$FATALITIES,
                        INJURIES = eventData$INJURIES)
  # Add a variable indicating if the event was pre-1970:
  costData$PRE1970 = eventData$PTIME$year < 70
```

The NOAA dataset stores property damage cost as a small number, **PROPDMG**, and an exponent, **PROPDMGEP**.
We now convert these pair into a single numeric value.
I was unable to find documentation on what the exponents "h", "+", "-", or "?" denoted, so the cost from these events has been removed by setting it to zero. 
The same goes for the **CROPDMG** and **CROPDMGEXP** variables.
```{r, cache = TRUE}
  # Add variables for the cost
  expToFactor <- function(E){
    Enum <- suppressWarnings(as.numeric(E))
    if(!is.na(Enum)){
      10^Enum
    } else if(E == ""){
      1
    } else {
    switch(E,
           "K" = 10^3, "k" = 10^3,
           "M" = 10^6, "m" = 10^6,
           "B" = 10^9, "b" = 10^9,
           0)
    }
  }
  # Conver the exponents into numbers we can actually multiply by!
  numExp <- (apply(as.matrix(as.character(data$PROPDMGEXP)), MARGIN = 1 , FUN = expToFactor))
  costData$PROPDMG <- eventData$PROPDMG * numExp
  numExp <- (apply(as.matrix(as.character(data$PROPDMGEXP)), MARGIN = 1 , FUN = expToFactor))
  costData$CROPDMG <- eventData$CROPDMG * numExp
  costData$TOTALDMG <- costData$PROPDMG + costData$CROPDMG
```

We now create an aggregate table of costs by events type, and by pre-9170 versus post-1970 events.
This latter choice is made because the latter dataset is more likely to be complete.
The earlier tables are removed to free up memory.

```{r, cache = TRUE}
  aData <- aggregate(cbind(INJURIES, FATALITIES, TOTALDMG) ~ PRE1970 + EVTYPE, data = costData, FUN = sum)
  rm(eventData)
  rm(costData)
```

## Results

### Population health harm of weather events

Due to the difficulty in assigning a relative weighting to injuries and deaths, we report seperately on each.
We only include events after 1970, since later dater is likely to be more complete.
Due to the large number of event types in the dataset we report only on the six causing the highest number of deaths and injuries.

```{r, fig.width = 480, fig.height = 480}
  fatalData <- aData[order(aData$FATALITIES, decreasing=TRUE), ]
  fatalData <- head(fatalData[fatalData$PRE1970 == FALSE, ], n=6)
  barplot(fatalData$FATALITIES, names.arg = fatalData$EVTYPE, cex.names=0.6,
          main = "Fatalities by event type", ylab = "Fatalities since 1970")
```

We can see that the most deadly event type is `r tolower(fatalData$EVTYPE[1])`, responsible for `r fatalData$FATALITIES[1]` deaths since 1970, followed by `r tolower(fatalData$EVTYPE[2])`, causing `r fatalData$FATALITIES[2]` deaths since 1970.


```{r, fig.width = 480, fig.height = 480}
  injData <- aData[order(aData$INJURIES, decreasing=TRUE), ]
  injData <- head(injData[injData$PRE1970 == FALSE, ], n=6)
  barplot(injData$INJURIES, names.arg = injData$EVTYPE, cex.names=0.6,
          main = "Injuries by event type", ylab = "Injuries since 1970")
```

We can see that the most non-fatal injuries were caused by `r tolower(injData$EVTYPE[1])`, responsible for `r injData$INJURIES[1]` injuries since 1970.
This is more than twice the number of non-fatal injuries from the next 5 sources put together.

### Financial harm of weather events

Again, we report data from 1970 onwards in an effort to improve completeness.

```{r, fig.width = 480, fig.height = 480}
  dmgData <- aData[order(aData$TOTALDMG, decreasing=TRUE), ]
  dmgData <- head(dmgData[dmgData$PRE1970 == FALSE, ], n=6)
  barplot(dmgData$TOTALDMG, names.arg = dmgData$EVTYPE, cex.names=0.5, axes=F,
          ylab = "Financial damage since 1970 (trillions (10^12) of dollars)",
          main = "Financial damage by event type")
  axis(2, at = c(0, 2e11, 4e11, 6e11, 8e11), labels = c("0", "0.2", "0.4", "0.6", "0.8"))
```

The leading two event types under the NOAA classification, hurricane and "hurricane / typhoon", are really the same physical phenomenon.
Between them, these account for a total of `r (dmgData$TOTALDMG[1] + dmgData$TOTALDMG[2])/1e12` trillion dollars of financial damage since 1970, dwarfing the next greatest source of damage, `r tolower(dmgData$EVTYPE[3])`, at `r dmgData$TOTALDMG[3]/1e9` billion dollars of damage since 1970.

## Conclusion

The main weather-realted causes of financial loss (by a very large margin) are hurricanes and typhoons.
However, these do not enter into the top six causes of fatalities or injuries, both of which are mainly due to tornadoes.
I would speculate that this is due to the much longer warning period preceding a hurricane making landfall, allowing evacuation of people, leaving only buildings vulnerable.
Tornadoes occur on a much shorter timescale and are not preceded by a warning in the same way.
